
# Activity 1

## Load packages

```{r message=FALSE, warning=FALSE}
library(sf)
library(lubridate)
library(mgcv)
library(rpart)
library(party)
library(leaflet)
library(patchwork)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Obtain a .gpx or .csv file for your movement data.

```{r message=FALSE, warning=FALSE, include=FALSE}
url <- "https://www.dropbox.com/scl/fi/mt4qyh9tmjjp1qyn7rq1f/Morning_Walk.gpx?rlkey=rr72rkduwgsorgpn2eafowpam&dl=1"

str(data)
```

## Plot/map your movement data. I would recommend using R and/or Google earth as I demonstrated in class.

```{r message=FALSE, warning=FALSE, include=FALSE}
# Read the GPX data
data <- st_read("Morning_Walk.gpx", layer = "track_points")
```

```{r}
# Plot the movement data
leaflet(data) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>% 
  addCircleMarkers(
    color = "darkorange", 
    fill = TRUE, fillColor = "darkorange", fillOpacity = 0.8, radius = 2,
    popup = ~paste("Time:", format(time, "%T")))
```

## Explore your movement data. For example, are there any unique features of your data (e.g., a large change in location)? Do your data contain location error? Really try to explore your data as best as possible using the plots/maps you made in 3.

```{r fig.height=4, fig.width=12, message=FALSE, warning=FALSE}
# Extract coordinates and convert to a data frame
coords <- as.data.frame(st_coordinates(data))

# Add elevation and time
coords$ele <- data$ele
coords$time <- data$time

# Convert time to seconds
coords$time <- as.numeric(difftime(coords$time, min(coords$time), units = "secs"))

coords <- coords %>% 
  rename(Longitude = X, 
         Latitude = Y,
         Elevation = ele,
         Time = time)
  
# Data for plotting
coords_plot <- coords %>% 
  pivot_longer(-Time, names_to = "Variable", values_to = "Value")

ggplot(coords_plot, aes(x = Time, y = Value, color = Variable)) +
  geom_point(show.legend = F) +
  facet_wrap(~Variable, scales = "free_y") +
  scale_color_brewer(palette = "Set1") + 
  theme_bw() +
  theme(text = element_text(size = 18))
```

## Fit a statistical or machine learning model to your movement data. Obtain predictions of your location on a fine time scale so that the estimates resemble a continuous trajectory.

```{r message=FALSE, warning=FALSE, }
# Fit model to longitude using time as a predictor
m1_long <- lm(Longitude ~ poly(Time, degree = 20, raw = T), data = coords)
m2_long <- gam(Longitude ~ s(Time, bs="gp", k=50), data = coords)

#summary(m1_long)
#summary(m2_long)
```

```{r message=FALSE, warning=FALSE}
# Fit model to latitude using time as a predictor
m1_lat <- lm(Latitude ~ poly(Time, degree = 20, raw = T), data = coords)
m2_lat <- gam(Latitude ~ s(Time, bs="gp", k=50), data = coords)

#summary(m1_lat)
#summary(m2_lat)
```

```{r message=FALSE, warning=FALSE}
# Fit model to elevation using time as a predictor
m1_elev <- lm(Elevation ~ poly(Time, degree = 20, raw = T), data = coords)
m2_elev <- gam(Elevation ~ s(Time, bs="gp", k=50), data = coords)

#summary(m1_elev)
#summary(m2_elev)
```

```{r message=FALSE, warning=FALSE}
# Estimate movement trajectory at a very fine temporal scale (every 1/2th of a second)
df.pred <- data.frame(Time = seq(0, as.numeric(max(coords$Time)), 0.5))

df.pred$lon.m1.hat = predict(m1_long, newdata = df.pred)
df.pred$lon.m2.hat = predict(m2_long, newdata = df.pred)

df.pred$lat.m1.hat = predict(m1_lat, newdata = df.pred)
df.pred$lat.m2.hat = predict(m2_lat, newdata = df.pred)

df.pred$elev.m1.hat = predict(m1_elev, newdata = df.pred)
df.pred$elev.m2.hat = predict(m2_elev, newdata = df.pred)
```

## Plot/map your estimated trajectory from 5. Explore your estimated trajectory as best as possible using the plots/maps. Note any unique features or shortcomings of your model.

```{r fig.height=4.5, fig.width=12, message=FALSE, warning=FALSE}
# Plot for Longitude
plt.long <- ggplot() +
  geom_point(data = coords, aes(x = Time, y = Longitude), size = 2, color = "grey") +
  geom_line(data = df.pred, aes(x = Time, y = lon.m1.hat), color = "navy", size = 1) +
  geom_line(data = df.pred, aes(x = Time, y = lon.m2.hat), color = "darkorange", size = 1) +
  labs(x = "Time", y = "Longitude") +
  theme_bw() +
  theme(text = element_text(size = 18))

# Plot for Latitude
plt.lat <- ggplot() +
  geom_point(data = coords, aes(x = Time, y = Latitude, color = "Observed data"), size = 2) +
  geom_line(data = df.pred, aes(x = Time, y = lat.m1.hat, color = "Polynomial prediction"), size = 1) +
  geom_line(data = df.pred, aes(x = Time, y = lat.m2.hat, color = "GAM prediction"), size = 1) +
  scale_color_manual(values = c("Polynomial prediction" = "navy", 
                                "GAM prediction" = "darkorange",
                                "Observed data" = "grey50")) +
  labs(x = "Time", y = "Latitude") +
  theme_bw() +
  theme(text = element_text(size = 18),
        legend.position = "top", 
        legend.direction = "vertical",
        legend.title = element_blank())

# Plot for Elevation
plt.elev <- ggplot() +
  geom_point(data = coords, aes(x = Time, y = Elevation), size = 2, color = "grey") +
  geom_line(data = df.pred, aes(x = Time, y = elev.m1.hat), color = "navy", size = 1) +
  geom_line(data = df.pred, aes(x = Time, y = elev.m2.hat), color = "darkorange", size = 1) +
  labs(x = "Time", y = "Elevation") +
  theme_bw() +
  theme(text = element_text(size = 18))

plt.long + plt.lat + plt.elev
```

```{r message=FALSE, warning=FALSE}
# Create data frame for plotting
df.pred.poly <- df.pred %>% 
  dplyr::select(Time, lon.m1.hat, lat.m1.hat) %>% 
  rename(Longitude = lon.m1.hat, Latitude = lat.m1.hat) %>% 
  mutate(Model = "Polynomial")

df.pred.cart <- df.pred %>% 
  dplyr::select(Time, lon.m2.hat, lat.m2.hat) %>% 
  rename(Longitude = lon.m2.hat, Latitude = lat.m2.hat) %>% 
  mutate(Model = "GAM")

df.pred.map <- df.pred.cart %>% rbind(df.pred.poly) %>% 
  rbind(coords %>% dplyr::select(-Elevation) %>% mutate(Model = "Observed"))
  
color.model <- colorFactor(palette = c("Polynomial" = "darkorange", "GAM" = "navy", "Observed" = "yellow"), 
                           domain = df.pred.map$Model)

# Plot the movement data
leaflet(df.pred.map) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>% 
  addCircleMarkers(
    ~Longitude,
    ~Latitude,
    fillColor = ~color.model(Model), 
    color = ~"grey50",
    radius = 3,
    weight = 0.5,
    fillOpacity = 1) %>%  
  addLegend(
    title = "Model",
    position = "topright",
    pal = color.model,
    values = ~Model, 
    opacity = 1)
```

## Estimate a feature or quantity of interest from your estimated trajectory (e.g., velocity, residence time, number of contacts, etc)

```{r message=FALSE, warning=FALSE}
# Calculate speed observed data
dist <- st_distance(data$geometry[1:809], data$geometry[2:810], by_element = T)

(sum(dist, na.rm = T)/1000)*.62 # looks okay (Strava = 0.74 mi)

speed <- (dist/as.numeric(diff(data$time)))*2.24 

plot(coords$Time[-1], speed, xlab="Time (seconds)",ylab="Velocity (miles per hour)", main = 'Observed')
```

```{r message=FALSE, warning=FALSE}
#Convert model coordinates to sf object
data.hat.m1 <- st_as_sf(df.pred, coords = c("lon.m1.hat", "lat.m1.hat"), 
                           crs = st_crs(data))

data.hat.m2 <- st_as_sf(df.pred, coords = c("lon.m2.hat", "lat.m2.hat"), 
                           crs = st_crs(data))
```


```{r message=FALSE, warning=FALSE}
# Calculate speed Polynomial
dist.hat.m1 <- st_distance(data.hat.m1$geometry[1:1618], data.hat.m1$geometry[2:1619], by_element = T)

(sum(dist.hat.m1)/1000)*.62

speed.hat.m1 <- (dist.hat.m1/as.numeric(diff(data.hat.m1$Time)))*2.24

plot(data.hat.m1$Time[-1], speed.hat.m1, xlab="Time (seconds)",ylab="Velocity (miles per hour)", main = 'Model: Polynomial')
```


```{r message=FALSE, warning=FALSE}
# Calculate speed GAM
dist.hat.m2 <- st_distance(data.hat.m2$geometry[1:1618], data.hat.m2$geometry[2:1619], by_element = T)

(sum(dist.hat.m2)/1000)*.62

speed.hat.m1 <- (dist.hat.m2/as.numeric(diff(data.hat.m2$Time)))*2.24

plot(data.hat.m2$Time[-1], speed.hat.m1, xlab="Time (seconds)",ylab="Velocity (miles per hour)", main = 'Model: GAM')
```








